
\documentclass{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{framed}

\hoffset=0pt
\oddsidemargin=15pt
\voffset=0pt
\marginparwidth = 2pt
\textheight = 642pt
\textwidth = 420pt


\begin{document}
	
	\tableofcontents
\begin{framed}
This is the draft document for in-session slides.
\end{framed}
\textbf{Pre-requisites}
\begin{itemize}
\item Downloading and installing R packages
\item What is a data frame
\item Principles of Tidy Data
\end{itemize}
\newpage




\section{dplyr : Grammar of data manipulation}
	\begin{itemize}
		\item dplyr  is mainly authored by Hadley Wickham and Romain Francois. It is designed to be intuitive and easy to learn, thereby making “doing things” in \texttt{R} more user friendly.
		\item dplyr is a new package which provides a set of tools for efficiently manipulating datasets in \texttt{R}.
		\item dplyr is the next iteration of plyr, focussing on only data frames. \item dplyr is faster, has a more consistent API and should be easier to use. 
	\end{itemize}
	
	
\subsection{dplyr : abstract by Hadley Wickham}

	\begin{framed}
	
	
	\noindent There are three key ideas that underlie dplyr:
	
	\begin{enumerate}
	\item Your time is important, so Romain Francois has written the key pieces in Rcpp to provide blazing fast performance. Performance will only get better over time, especially once we figure out the best way to make the most of multiple processors. 

	\item Tabular data is tabular data regardless of where it lives, so you should use the same functions to work with it. 
	
	With dplyr, anything you can do to a local data frame you can also do to a remote database table. PostgreSQL, MySQL, SQLite and Google bigquery support is built-in; adding a new backend is a matter of implementing a handful of S3 methods. 

	\item The bottleneck in most data analyses is the time it takes for you to figure out what to do with your data, and dplyr makes this easier by having individual functions that correspond to the most common operations  group\_by, summarise,mutate, filter, select and arrange). Each function does one only thing, but does it well.
	\end{enumerate}
	Author: Hadley Wickham
    \end{framed}


\subsection{Working with dplyr} \textbf{dplyr} focussed on tools for working with data frames (hence the d in the name). \textbf{dplyr} has three main goals:


\begin{itemize}
\item Identify the most important data manipulation tools needed for data analysis and make them easy to use from \texttt{R}.

\item Provide very fast performance for in-memory data by writing key pieces in C++.

\item Use the same interface to work with data no matter where it's stored, whether in a data frame, a data table or database.
\end{itemize}
\subsection{Installing dplyr}
You can install the latest released version from CRAN with the code below.
You can also install and load the data packages used in most examples: 
\begin{framed}
\begin{verbatim}
install.packages("dplyr")
install.packages(c("nycflights13", "Lahman"))

library(dplyr) # for functions
library(nycflights13) # for data
\end{verbatim}
\end{framed}

\subsection{Tidy Data}
To make the most of dplyr, Hadley Wickham recommends that you familiarise yourself with the \textbf{principles of tidy data}. This will help you get your data into a form that works well with \textbf{dplyr}, \textbf{ggplot2} and \texttt{R}'s many modelling functions.\\
\bigskip

\begin{framed}
\noindent Three Principles from Hadley Wickham's paper
	\begin{itemize}
		\item[1.] Each variable forms a column, 
		\item[2.] Each observation forms a row, 
		\item[3.] Each table/file stores data about one kind of observation.
	\end{itemize}
\end{framed}
\noindent \textbf{Remark:}  The paper ``\textit{\textbf{Tidy data}}" by Hadley Wickham (RStudio) can be downloaded from 
 \begin{verbatim}
http://vita.had.co.nz/papers/tidy-data.pdf
 \end{verbatim}

%=================================================================== %

\subsection{Key data structures}

The key object in \textbf{dplyr} is a \texttt{tbl}, a representation of a tabular data structure. Currently dplyr supports:

\begin{itemize}
\item data frames - the  most commonly encountered R data structure. 
\item data tables - a data structure that is designed for intensive data analysis.
\end{itemize}

\noindent For this class, We will concentrate on \textbf{dplyr} exercises with data frames mostly. However we would advise you to try and learn more about working with data tables in the future.\\
\bigskip

\noindent For advanced users, \textbf{dplyr} also supports the following databases: \textit{SQLite, PostgreSQL, Redshift, MySQL/MariaDB, Bigquery, MonetDB} and data cubes with arrays (partial implementation). We will not cover those topics in this workshop.

%================================================================================ %
\newpage
\section{Worked Examples}

\subsection*{New York Flights Exercise}

To explore the basic data manipulation verbs of dplyr, we’ll start with the built in \textit{nycflights13} data frame. This dataset comes from the US Bureau of Transporation Statistics contains information on flights that departed from New York City in 2013. 


\begin{framed}
\begin{verbatim}
library(dplyr) # for functions
library(nycflights13) # for data
data(flights)
\end{verbatim}
\end{framed}

\subsubsection*{Preliminary Exercises}
Using simple data inspection functions, answer the following questions.
\begin{itemize}
\item How many columns (i.e. variables) are there in this data set?
\item How many rows (i.e. cases) are there?
\item Which column has the most missing values?
\item What are names of the last three variables?
\item What type of data object is ``flights"?
\end{itemize}

\begin{framed}
\begin{verbatim}
#>    year month day dep_time dep_delay arr_time arr_delay carrier tailnum
#> 1  2013     1   1      517         2      830        11      UA  N14228
#> 2  2013     1   1      533         4      850        20      UA  N24211
#> 3  2013     1   1      542         2      923        33      AA  N619AA
#> 4  2013     1   1      544        -1     1004       -18      B6  N804JB
#> ..  ...   ... ...      ...       ...      ...       ...     ...     ...
\end{verbatim}
\end{framed}
\subsection{Converting to tbl}
dplyr can work with data frames as is, but if you’re dealing with large data, it’s worthwhile to convert them to a \texttt{tbl\_df}: this is a wrapper around a data frame that won’t accidentally print a lot of data to the screen.




\subsection{Filter rows with \texttt{filter()}}

\texttt{filter()} allows you to select a subset of the rows of a data frame. The first argument is the name of the data frame, and the second and subsequent are filtering expressions evaluated in the context of that data frame:
For example, we can select all flights on January 1st with:

\begin{framed}
\begin{verbatim}
filter(flights, month == 1, day == 1)
\end{verbatim}
\end{framed}



\subsection{ \texttt{group\_by()}}

The \texttt{tbl} also comes in a grouped variant which allows you to easily perform operations "by group". Suppose we wish to group by carrier.

\begin{verbatim}
carriers_df  <- group_by(flights, carrier)
\end{verbatim}

\end{document}